Formalization, Mechanization and Automation of G\"{o}del's Proof of God's Existence\thanks{This work has been supported by the German Research Foundation under grant BE2501/9-1.}

### 2013-08-19 Abstract/GodProof-Abstract.tex
improved text of abstract

\usepackage{url,amsmath,amssymb}

% Logical symbols
\newcommand{\imp}{\rightarrow}
\newcommand{\biimp}{\leftrightarrow}
\newcommand{\all}{\forall}
\newcommand{\ex}{\exists}
\newcommand{\seq}{\vdash}
\newcommand{\nec}{\Box} % necessarily
\newcommand{\pos}{\Diamond} % possibly
\newcommand{\ess}[2]{#1 \ \mathit{ess} \ #2}

\title{ Formalization, Mechanization and Automation % of Variants
  of G\"{o}del's Proof of God's Existence\thanks{This work has been
    supported by the German Research Foundation under grant
    BE2501/9-1. Bruno ????}  }

@@

\institute{
  Dahlem Center for Intelligent Systems, Freie Universit\"{a}t Berlin, Germany\\
  \email{c.benzmueller@gmail.com}
  \and 
  Theory and Logic Group, Vienna University of Technology, Austria \\
  \email{bruno@logic.at}
}

@@

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology.  G\"{o}del's proof \cite{Goedel1970} is a modern culmination of
this tradition, following particularly the footsteps of Leibniz.
%
G\"{o}del defines God as a being who possesses all \emph{positive} properties.
He does not extensively discuss what positive properties are, 
but instead he states a few reasonable but debatable axioms that they should satisfy.
Various slightly different versions of axioms and definitions have been considered by G\"{o}del and by several philosophers who commented on his proof (e.g. \cite{Scott,Sobel,AndersonGettings,Fitting,Adams,ContemporaryBibliography}). Our formalization employs the following axioms (A*) and definitions (D*):\footnote{
(A1), (A2), (A5), (D1), (D3) are logically equivalent to, respectively, axioms 5, 2 and 4 and definitions 1 and 3 in G\"odel's manuscript \cite{Goedel1970}. (A3) was introduced by Scott \cite{Scott} and could be derived from G\"odel's axiom 1 and (D1) in a logic with infinitary conjunction. (A4) is a weaker form of G\"odel's axiom 3. (D2) has an extra conjunct lacking in G\"odel's definition 2; this lack is believed to have been an oversight by G\"odel \cite{Hazen}.
}
\marginpar{\scriptsize Chris: The reference to Scott's version is kind of confusing, we need to verify this source.}
\marginpar{\scriptsize Bruno: I think the numbers (A1, A2,.., D1,..D3, L1,..,L3,T) should be in the left with a description environment, as they were before. I think it looked nicer like that. If you agree, I can change back.}
\allowdisplaybreaks[1] 
\begin{align}
& \text{Any property necessarily implied by a positive property is positive.} \notag \\
& \quad 
\all \varphi. \all \psi.[(P(\varphi) \wedge \nec \all x.[\varphi(x) \imp \psi(x)]) \imp P(\psi)] \tag{A1} \\
& \text{A property is positive if and only if its negation is not positive.} \notag \\
& \quad 
\all \varphi. [P(\neg \varphi) \biimp \neg P(\varphi)]
\tag{A2} \\
& \text{The property of being God-like is positive.} \notag \\
& \quad 
P(G) 
\tag{A3} \\
& \text{Positive properties are necessarily positive.} \notag \\
& \quad 
\all \varphi.[P(\varphi) \to \Box \; P(\varphi)]
\tag{A4} \\
& \text{Necessary existence is a positive property.} \notag \\
& \quad 
P(E)
\tag{A5} \\
& \text{x is God-like if and only if x has every positive property.} \notag \\
& \quad 
G(x) \biimp \forall \varphi. [P(\varphi) \to \varphi(x)]
\tag{D1} \\ 
& \parbox[c][2em][c]{0.8\textwidth}{A property P is an essence of x if and only if P is a property of x and every property Q that x has is necessarily implied by P.}\notag \\
& \quad 
\ess{\varphi}{x} \biimp \varphi(x) \wedge \all \psi. (\psi(x) \imp \nec \all x. (\varphi(x) \imp \psi(x)))
\tag{D2} \\
& \parbox[c][2em][c]{0.8\textwidth}{x necessarily exists if and only if every essence of x is necessarily exemplified.} \notag \\
& \quad 
E(x) \biimp \all \varphi.[\ess{\varphi}{x} \imp \nec \ex x.\varphi(x)] 
\tag{D3} 
\end{align}

\noindent
From these axioms and definitions we then infer:
\begin{align}
& \text{Positive properties are possibly exemplified.} \notag \\
& \quad 
\all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]
\tag{L1} \\
& \text{Possibly God exists.} \notag \\
& \quad 
\pos \ex x. G(x)
\tag{L2} \\
& \parbox[c][2em][c]{0.8\textwidth}{If x is God-like, then the property of being God-like is an essence of x.} \notag \\
& \quad 
\all y.[G(y) \imp \ess{G}{y}]
\tag{L3} \\
& \text{Necessarily God exists.} \notag \\
& \quad 
\nec \ex x. G(x)
\tag{T}
\end{align}

The above variant of G\"{o}del's proof has now been 
constructed for the first-time
with an unprecedent degree of detail and formality; cf.~\cite{FormalTheologyRepository}. The following has been done (and in this order):
\begin{itemize}
\item A detailed natural deduction proof.\marginpar{\scriptsize Chris: We need to make sure that this ND proof is based on a sound calculus.}
%
\item A formalization of the axioms, definitions and theorems in the TPTP THF syntax \cite{J22}.
%
\item Automatic verification of the consistency of the axioms and 
definitions with Nitpick \cite{Nitpick}.
%
\item Automatic demonstration of the theorems with the provers LEO-II \cite{LEO-II} and Satallax \cite{Satallax}.

@@

G\"{o}del's proof is challenging to formalize and verify because it
requires an expressive logical language with modal operators
(\emph{possibly} and \emph{necessarily}) and with
quantififiers for individuals and sets of individuals (properties).  Our computer-assisted formalizations rely on an
embedding of the modal logic S5 into classical higher-order logic with
Henkin semantics \cite{J23,B9}. The formalization is thus essentially
done in classical higher-order logic where quantified S5 is emulated.

@@

Future work includes an extensive study of other formalizations of
ontological arguments with our machinery. Variations of the
distinctive features of the base logic S5 (e.g. non-rigid symbols, 
varying domains, etc.) are enabled in these studies due
to the flexibility of the embeddings based approach.

\bibliographystyle{plain}
\bibliography{Bibliography}

>>>>

\usepackage{url,amsmath,amssymb,a4wide}

\newcommand{\imp}{\rightarrow}
\newcommand{\biimp}{\leftrightarrow}
\newcommand{\all}{\forall}
\newcommand{\ex}{\exists}
\newcommand{\seq}{\vdash}
\newcommand{\nec}{\Box} % necessarily
\newcommand{\pos}{\Diamond} % possibly
\newcommand{\ess}[2]{#1 \ \mathit{Ess.} \ #2}
\newcommand{\NE}{\mathit{NE}}

\title{ Formalization, Mechanization and Automation % of Variants
  of G\"{o}del's Proof of God's Existence\thanks{This work has been
    supported by the German Research Foundation under grant
    BE2501/9-1.}  }

@@

\institute{
  Dahlem Center for Intelligent Sstems, Freie Universit\"{a}t Berlin, Germany\\
  \email{c.benzmueller@gmail.com}
  \and 
  Theory and Logic Group, Vienna University of Technology, Austria \\
  \email{bruno@logic.at}
}

@@

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology.  G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern culmination of
this tradition, following particularly the footsteps of Leibniz.
%
G\"{o}del defines God as a being who possesses all \emph{positive} properties.
He does not extensively discuss what positive properties are, 
but instead he states a few reasonable (but debatable) axioms that they should satisfy.
Various slightly different versions of axioms and definitions have
been considered by G\"{o}del and by several philosophers who commented
on his proof
(cf. \cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).

Dana Scott's version of G\"odel's proof \cite{ScottNotes} employs the
following axioms (\textbf{A}), definitions (\textbf{D}), corollaries
(\textbf{C}) and theorems (\textbf{T}), and it proceeds in the
following order:\footnote{ (A1), (A2), (A5), (D1), (D3) are logically
  equivalent to, respectively, axioms 5, 2 and 4 and definitions 1 and
  3 in G\"odel's notes \cite{Goedel1970,GoedellNotes}. (A3) was introduced by
  Scott \cite{ScottNotes} and could be derived from G\"odel's axiom 1 and
  (D1) in a logic with infinitary conjunction. (A4) is a weaker form
  of G\"odel's axiom 3. (D2) has an extra conjunct lacking in
  G\"odel's definition 2; this lack is believed to have been an
  oversight by G\"odel \cite{Hazen}. }
% \marginpar{\scriptsize Chris: The reference to Scott's version is kind of confusing, we need to verify this source.}
% \marginpar{\scriptsize Bruno: I think the numbers (A1, A2,..,
% D1,..D3, L1,..,L3,T) should be in the left with a description
% environment, as they were before. I think it looked nicer
% before. Can I change it back?}

\allowdisplaybreaks[1] 
\begin{description}
\item[A1] Either the property or its negation is positive, but not
  both:  \hfill 
  $\all \phi. [P(\neg \phi) \biimp \neg P(\phi)]$
\item[A2] A property is positive if it necessarily  contains a
  positive property:  \\ \phantom{b} \hfill 
  $\all \phi. \all \psi.[(P(\phi) \wedge \nec \all x.[\phi(x)
  \imp \psi(x)]) \imp P(\psi)]$ 
\item[T1] Positive properties are possibly exemplified. \hfill $\all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]$
\item[D1] x is God-like if it possesses all positive properties: \hfill
  $G(x) \biimp \forall \phi. [P(\phi) \to \phi(x)]$
\item[A3]  The property of being God-like is positive: \hfill   $P(G)$
\item[C\phantom{1}] Possibly, God exists: \hfill $\pos \ex x. G(x)$
\item[A4]  Being a positive property is logical, hence,
  necessary: \hfill 
  $\all \phi.[P(\phi) \to \Box \; P(\phi)]$
\item[D2] $\phi$ is the essence of $x$ iff $x$ has $\phi$ and this
  property is necessarily minimal.:\\
  \phantom{b} \hfill $\ess{\phi}{x} \biimp \phi(x) \wedge \all
  \psi. (\psi(x) \imp \nec \all y. (\phi(y) \imp \psi(y)))$
\item[T2]  The property of being God-like is an essence of any
  God-like being: \hfill $\all y.[G(y) \imp \ess{G}{y}]$ 
\item[D3] $\NE(x)$ means that $x$ necessarily exists if it has an
  essential property: \\
  \phantom{b} \hfill $\NE(x) \biimp \all \phi.[\ess{\phi}{x} \imp \nec \ex x.\phi(x)]$
\item[A5] Necessary existence is a positive property: \hfill $P(\NE)$
\item[T3] Necessarily, God exists: \hfill $\nec \ex x. G(x)$ 
\end{description}

% \begin{align}
% & \text{Any property necessarily implied by a positive property is positive.} \notag \\
% & \quad 
% \all \varphi.\all \psi.[(P(\varphi) \wedge \nec \all x.[\varphi(x) \imp \psi(x)]) \imp P(\psi)] \tag{A1} \\
% & \text{A property is positive if and only if its negation is not positive.} \notag \\
% & \quad 
% \all \varphi. [P(\neg \varphi) \biimp \neg P(\varphi)]
% \tag{A2} \\
% & \text{The property of being God-like is positive.} \notag \\
% & \quad 
% P(G) 
% \tag{A3} \\
% & \text{Positive properties are necessarily positive.} \notag \\
% & \quad 
% \all \varphi.[P(\varphi) \to \Box \; P(\varphi)]
% \tag{A4} \\
% & \text{Necessary existence is a positive property.} \notag \\
% & \quad 
% P(E)
% \tag{A5} \\
% & \text{Being God-like is having every positive property.} \notag \\
% & \quad 
% G(x) \biimp \forall \varphi. [P(\varphi) \to \varphi(x)]
% \tag{D1} \\ 
% & \parbox[c][2em][c]{0.8\textwidth}{A property is an essence of an individual if and only if the individual has it and every property  that the individual has is necessarily implied by it.}\notag \\
% & \quad 
% \ess{\varphi}{x} \biimp \varphi(x) \wedge \all \psi. (\psi(x) \imp \nec \all y. (\varphi(y) \imp \psi(y)))
% \tag{D2} \\
% & \parbox[c][2em][c]{0.8\textwidth}{An individual necessarily exists if and only if all its essences are necessarily exemplified.} \notag \\
% & \quad 
% E(x) \biimp \all \varphi.[\ess{\varphi}{x} \imp \nec \ex x.\varphi(x)] 
% \tag{D3} 
% \end{align}

% \noindent
% From these axioms and definitions we then infer:
% \begin{align}
% & \text{Positive properties are possibly exemplified.} \notag \\
% & \quad 
% \all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]
% \tag{L1} \\
% & \text{Possibly, God exists.} \notag \\
% & \quad 
% \pos \ex x. G(x)
% \tag{L2} \\
% & \parbox[c][2em][c]{0.8\textwidth}{The property of being God-like is an essence of any God-like being.} \notag \\
% & \quad 
% \all y.[G(y) \imp \ess{G}{y}]
% \tag{L3} \\
% & \text{Necessarily, God exists.} \notag \\
% & \quad 
% \nec \ex x. G(x)
% \tag{T}
% \end{align}
% \marginpar{\scriptsize Bruno: I think it might be possible to improve the natural language statements further. After our improvements converge here, we should update the statements in all formalizations (THF, Coq, Isabelle, ND).}

The above version of G\"{o}del's proof has now been 
analysed for the first-time
with an unprecedent degree of detail and formality; cf.~\cite{FormalTheologyRepository}. The following has been done (and in this order):
\begin{itemize}
\item A detailed natural deduction proof.
%
\item A formalization of the axioms, definitions and theorems in the TPTP THF syntax \cite{J22}.
%
\item Automatic verification of the consistency of the axioms and 
definitions with Nitpick \cite{Nitpick}.
%
\item Automatic demonstration of the theorems with the provers LEO-II \cite{LEO-II} and Satallax \cite{Satallax}.

@@

G\"{o}del's proof is challenging to formalize and verify because it
requires an expressive logical language with modal operators
(\emph{possibly} and \emph{necessarily}) and with
quantifiers for individuals and properties.  Our computer-assisted formalizations rely on an
embedding of the modal logic into classical higher-order logic with
Henkin semantics \cite{J23,B9}. The formalization is thus essentially
done in classical higher-order logic where  quantified modal logic is
emulated.

In our ongoing computer-assisted study of G\"odel's proof we have
obtained the following results:
\begin{itemize}
\item Logic S5 is not needed for obtaining the results, logic KB is
  sufficient.
\item For proving theorem T1, only the left to right direction of
  axiom A1 is needed. However, the backward direction of A1 is
  required for showing T2.
\item The first conjunct $\phi(x)$ in D2, introduced by Scott, seems important; without it the set of axioms and definitions becomesx inconsistent.
\end{itemize}

@@

Future work includes a substantial widening of our computer-assisted
study of ontological arguments. Due to the flexibility of the
embeddings based approach, variations of the distinctive features of
the base logic (e.g. non-rigid symbols, varying domains, etc.)  will
be enabled in these studies

%\footnotesize
\bibliographystyle{plain}
\bibliography{Bibliography}
